Updated Constraints on Self-Interacting Dark Matter from Supernova 1987A

### 2017-01-07 draft.tex
Made some further changes through section 3.

@@

The quantity of interest is the rate of energy emission through dark gauge bosons. 
From the spin-summed, squared amplitudes described in the previous subsection, 
the energy emission rate is obtained by integrating over the phase space, 
and adding a factor of the energy of the emitted particle. To be specific, the 
energy emission rate per unit volume is 
%
\begin{widetext}
\beq
\label{eq:rate1}
Q_i = (2\pi)^4 \int\, E_{\mathrm{A}'} \, \sum_{s_1,s_2}\, \vert \mathcal{M}_i \vert^2 f(p_1) f(p_2)\delta^{(4)}(p_1+p_2-p_3-p_4-\qa)\, d\Pi,
\eeq
\end{widetext}
%
where $d\Pi$ is the Lorentz-invariant phase space, 
$ E_{\mathrm{A}'}$ is the energy of the emitted $A'$ boson, 
$f(p)$ are the phase-space densities of the incoming nucleons, and 
the index $i$ refers to either the pp or pn processes. 
The nucleons in the core are comfortably non-degenerate and non-relativistic, 
so the Pauli blocking factor is omitted from Eq.~(\ref{eq:rate1}) and we take 
all nucleons to have a Maxwell-Boltzmann momentum distribution, so 
\beq
f(p) =  \frac{n_b}{2} \left( \frac{2 \pi}{M_N T} \right)^{3/2} \exp \left(-\frac{p^2}{2 M_N T} \right).
\eeq
%
We chose a baryon number density of $n_{\mathrm{b}} \approx 1.8 \times 10^{38}\, \mathrm{cm}^{-3}$ and a 
core supernova temperature of $T = 30\, \mathrm{MeV}$, both of which are typical choices and thought to 
be representative of supernova cores. \arz{Need to put a citation here for the properties of supernova cores. 
I can probably find something in the literature, but please put something here if you have a citation.}

@@

To obtain the dark gauge boson luminosity from $Q_{\mathrm{pp}}$ and $Q_{\mathrm{pn}}$, 
we assumed that $A'$ production takes place in a stellar core of radius $\sim 1$~km, so that 
the total luminosity of $A'$ is 
%
\begin{equation}
L_{\mathrm{A}'} = V (Q_{\mathrm{pp}} + Q_{\mathrm{pn}})
\end{equation}
%
where $V$ is the volume of the sphere. Clearly, $L_{\mathrm{A}'}$ is proportional to $\varepsilon^2$, 
so that writing $L_{\mathrm{A}'} = \varepsilon^2 I_{\mathrm{A}'}(m_{\mathrm{A}'},T)$ and 
$L_{\nu}$ as the neutrino luminosity, we can write the constraint as 
%
\begin{equation}
\varepsilon \lesssim \sqrt{\frac{L_{\nu}}{I_{\mathrm{A}'}(m_{\mathrm{A}'}, T)}}.
\end{equation}
%
Following SOMEBODAY \arz{We need a citation here.}, we show a constraint taking 
$L_{\nu} \approx 10^{53}\, \mathrm{erg/s} \approx 4 \times 10^{37}\, \mathrm{MeV}^2$.

%---------------------------------- DECAY LIMIT
\subsection{Decay limit}
The constraint from the streaming limit above provides an upper bound on the allowed coupling (or lower bound on the excluded coupling) by considering the production on the $A'$. Naively it might be expected that this suffices, in that all higher couplings are excluded. However, in order to function as a cooling channel, enough of the produced dark gauge bosons must escape the supernova. As the coupling increases, so do processes that prevent the escape, and so the excluded region has an upper bound, above which the luminosity again drops below the neutrino luminosity. The first such limit may be found by considering decay of the dark bosons into Standard Model particles, and assuming that these SM particles are trapped in the supernova core and contribute nothing to the cooling. The dark boson has a typical lifetime of 
\beq
l = \frac{3 E_{A}}{N_{eff} m_A^2 \varepsilon^2},
\eeq
and so the fraction escaping the supernova before decaying is given by 
\beq
e^{r_{decay}/l} = e^{r_{decay} N_{eff} m_A^2 \varepsilon^2/(3 E_A)}.
\eeq
To take this into account, the above exponential factor is simply appended to the phase space integrand, and the calculation then proceeds as before. The limit is derived from the same equation, with the only complication being the fact that $ I_A $ is now a function of $ \varepsilon $, in addition to $ m_A, T $. This makes the numerical calculation slightly more complicated, but otherwise changes nothing of importance, with now the constraint a transcendental equation,  
\beq
\varepsilon \le \sqrt{\frac{4.1 \times 10^{37} MeV^2}{I_A(m_A, T, \varepsilon)}}.
\eeq

%---------------------------------- TRAPPING	
\subsection{Trapping limit}
The second constraint that produces an upper bound on the excluded region comes from considering trapping of dark bosons within the supernova. With a large enough coupling, the new particles will thermalize and then will be emitted from a spherical shell.  In this case the luminosity is given simply by the Steffan-Boltzmann law
\beq
\mathcal{L}_t  = 4\pi r^2 T_A^4 \sigma,
\eeq
where $ r $ is now the radius of the emitting shell and $ T_A $ its temperature. We can estimate $ r = 10$ km, since the density of the supernova drops drastically around that point. After taking that value for r, the bound on the luminosity translates into a bound on $ T_A $ 
\beq
T_A \le 9.586\rm\ MeV.
\eeq
That bound can then be translated into the desired bound on the coupling as a function of mass by assuming that the particles are emitted from an optical depth $ \tau = 2/3 $, and finding the temperature that corresponds to that optical depth. This is a somewhat involved calculation.
First, one needs a model for the density and temperature in the supernova. Following \cite{dent_etal12}, we assume 
\bea
\rho &=& \rho_p \left(\frac{R}{r}\right)^n,\\
T &=& T_R \left(\frac{\rho(r)}{\rho_R}\right)^{1/3},
\eea
with $  \rho_p  = 3\times10^{14}\ {\rm g/cm}^3, T_R = 30$ MeV, and taking $ n = 5 $. The optical depth is given by 
\beq
\tau = \int_{r_x}^{\infty} \kappa \rho dr,
\eeq
where $ \kappa$ is the opacity.  To find the opacity, we start from the reduced mean Rosseland opacity 
\beq
\frac{1}{\kappa \rho} = \int_{m_x}^{\infty} \frac{15}{4 \pi^4 T^5} \frac{E_A^2 e^{E_A/T} \sqrt{E_A^2 - m_A^2}}{(e^{E_A/T}-1)^2} l_A dE_A,
\eeq
where $ l_A $ is the mean free path.

@@

@@

\begin{figure}[th]
	\includegraphics[width=9cm]{endtoend.png}
	\caption{A comparison of the final result for the excluded region with the result from Ref.~\cite{dent_etal12}. The excluded region is clearly significantly less restrictive, and opens up the region of parameter space for $\varepsilon \approx  1e-8-9 $.
\arz{Cameron, some similar comments as for Figure 1. Can we write numbers as "$10^{-9}$" rather 
than "1e-09" and so on? Can we make the axis labels and tick mark labels larger?}	
	}
\end{figure}

>>>>

\newcommand{\dd}{\mathrm{d}}

@@

The quantity of interest is the rate of energy emission through dark gauge bosons. 
From the spin-summed, squared amplitudes described in the previous subsection, 
the energy emission rate is obtained by integrating over the phase space, 
and adding a factor of the energy of the emitted particle. To be specific, the 
energy emission rate per unit volume is 
%
\begin{widetext}
\beq
\label{eq:rate1}
Q_i = (2\pi)^4 \int\, E_{\mathrm{A}'} \, \sum_{s_1,s_2}\, \vert \mathcal{M}_i \vert^2 f(p_1) f(p_2)\delta^{(4)}(p_1+p_2-p_3-p_4-\qa)\, \dd \Pi,
\eeq
\end{widetext}
%
where 
%
\begin{equation}
d\Pi = \frac{\dd^3 \vec{q}_{\mathrm{A}'}}{(2\pi)^3 2E_{\mathrm{A}'}}\, \prod_{i=1}^{4}\, \frac{\dd^3 \vec{p}_i}{(2\pi^3) 2E_1}
\end{equation} 
is the Lorentz-invariant phase space interval, 
$ E_{\mathrm{A}'}$ is the energy of the emitted $A'$ boson, 
$f(p)$ are the phase-space densities of the incoming nucleons, and 
the index $i$ refers to either the pp or pn processes. 
The nucleons in the core are comfortably non-degenerate and non-relativistic, 
so the Pauli blocking factor is omitted from Eq.~(\ref{eq:rate1}) and we take 
all nucleons to have a Maxwell-Boltzmann momentum distribution, so 
\beq
f(p) =  \frac{n_b}{2} \left( \frac{2 \pi}{M_N T} \right)^{3/2} \exp \left(-\frac{p^2}{2 M_N T} \right).
\eeq
%
We chose a baryon number density of $n_{\mathrm{b}} \approx 1.8 \times 10^{38}\, \mathrm{cm}^{-3}$ and a 
core supernova temperature of $T = 30\, \mathrm{MeV}$, both of which are typical choices and thought to 
be representative of supernova cores. \arz{Need to put a citation here for the properties of supernova cores. 
I can probably find something in the literature, but please put something here if you have a citation.}

@@

To obtain the dark gauge boson luminosity from $Q_{\mathrm{pp}}$ and $Q_{\mathrm{pn}}$, 
we assumed that $A'$ production takes place in a stellar core of radius $\sim 1$~km, so that 
the total luminosity of $A'$ is 
%
\begin{equation}
L_{\mathrm{A}'} = V (Q_{\mathrm{pp}} + Q_{\mathrm{pn}})
\end{equation}
%
where $V$ is the volume of the sphere. Clearly, $L_{\mathrm{A}'}$ is proportional to $\varepsilon^2$, 
so that writing $L_{\mathrm{A}'} = \varepsilon^2 I_{\mathrm{A}'}(m_{\mathrm{A}'},T)$ and 
$L_{\nu}$ as the neutrino luminosity, we can write the constraint as 
%
\begin{equation}
\label{eq:bound1}
\varepsilon \lesssim \sqrt{\frac{L_{\nu}}{I_{\mathrm{A}'}(m_{\mathrm{A}'}, T)}}.
\end{equation}
%
Following SOMEBODY \arz{We need a citation here.}, we derive a constraint taking 
$L_{\nu} \approx 10^{53}\, \mathrm{erg/s} \approx 4 \times 10^{37}\, \mathrm{MeV}^2$.

The constraint derived in this manner from Eq.~(\ref{eq:bound1}) sets the lower limit on the 
exclusion band shown in Figure~\ref{fig:bound1}, which we will discuss in more detail 
below. If all of the $A'$ produced in the core left the supernova freely, the bound of values of 
$\varepsilon$ higher than those given by Eq.~(\ref{eq:bound1}) would be ruled out. However, 
as we have already mentioned, $\varepsilon$ can become sufficiently large that only a negligible 
amount of energy actually leaves the supernova core because of either $A'$ decays or scattering. 
These considerations place an upper limit on the values of $\varepsilon$ for which this constraint 
applies, and we discuss these effects in the next two subsections.

%---------------------------------- DECAY LIMIT
\subsection{The Decay Limit}

The constraint from the streaming limit above provides an upper bound on the allowed coupling 
(or lower bound on the excluded coupling) from the production on $A'$ bosons. 
In order to function as an effective cooling channel, enough of the dark gauge bosons 
must escape the supernova. As the coupling increases, so do processes that prevent this escape, 
and so the excluded region has an upper bound, above which the luminosity again drops below the 
neutrino luminosity. The first such limit may be found by considering decay of the dark bosons 
into Standard Model particles. Standard model particles will scatter and thermalize 
on a timescale much shorter than the timescale for the evolution of the core, so decays 
contribute nothing to the cooling.

The dark boson has a typical lifetime of 
\arz{Cameron, try to rephrase this. This is not the lifetime, but rather the 
mean distance to a decay.}
%
\beq
l = \frac{3 E_{A}}{N_{eff} m_A^2 \varepsilon^2},
\eeq
and so the fraction escaping the supernova before decaying is given by 
\beq
e^{r_{decay}/l} = e^{r_{decay} N_{eff} m_A^2 \varepsilon^2/(3 E_A)}.
\eeq
%
To account for the suppression of gauge boson luminosity due to decays we simply multiply the phase space 
integrand in Eq.~(\ref{eq:rate1}) by the exponential suppression factor, after which the calculation proceeds 
as in the previous subsection. The limit is derived in the same way, 
with the additional complication that $I_{\mathrm{A}'}$ is now a function of $\varepsilon $, in addition to $m_{\mathrm{A}'}$ and $T$. 
The constraint Eq.~(\ref{eq:bound1}), therefore, becomes a transcendental equation that must be solved numerically.

The luminosity in $A'$ will be an increasing function of $\varepsilon$ until decays suppress the gauge boson luminosity, 
at which point $L_{\mathrm{A}'}$ becomes a rapidly decreasing function of $\varepsilon$. Therefore, the excluded values of 
$\varepsilon$ at a given mass will generally have a lower bound set by the calculations of the previous subsection, and an 
upper bound set by decays.

%---------------------------------- TRAPPING	
\subsection{Trapping limit}

\arz{This section needs some work as it is a bit hard to follow.}
The second effect that produces an upper bound on the excluded region 
comes from considering trapping of dark bosons within the supernova. 
With a large enough coupling, the dark photons will thermalize and then 
will be emitted from an approximately spherical ``dark photosphere" at the radial position 
where the $A'$ mean free path becomes larger than the typical size of the supernova core. 
In this case the luminosity is given simply by Stefan's law, 
\beq
\mathcal{L}_t  = 4\pi r^2 T_A^4 \sigma,
\eeq
where $r$ is now the radius of the emitting shell and $T_A$ its temperature. 
We estimate the radius of this dark photosphere as $r=10$ km, 
because the density of the supernova drops drastically around that point. 
After taking that value for r, the bound on the luminosity translates into a bound on $ T_A $ 
\beq
T_A \le 9.586\rm\ MeV.
\eeq
That bound can then be translated into the desired bound on the coupling as a function of mass by assuming that the particles are emitted from an optical depth $ \tau = 2/3 $, and finding the temperature that corresponds to that optical depth. This is a somewhat involved calculation.
First, one needs a model for the density and temperature in the supernova. Following \cite{dent_etal12}, we assume 
\bea
\rho &=& \rho_p \left(\frac{R}{r}\right)^n,\\
T &=& T_R \left(\frac{\rho(r)}{\rho_R}\right)^{1/3},
\eea
with $  \rho_p  = 3\times10^{14}\ {\rm g/cm}^3, T_R = 30$ MeV, and taking $ n = 5 $. The optical depth is given by 
\beq
\tau = \int_{r_x}^{\infty} \kappa \rho dr,
\eeq
where $ \kappa$ is the opacity.  To find the opacity, we start from the reduced mean Rosseland opacity 
\beq
\frac{1}{\kappa \rho} = \int_{m_x}^{\infty} \frac{15}{4 \pi^4 T^5} \frac{E_A^2 e^{E_A/T} \sqrt{E_A^2 - m_A^2}}{(e^{E_A/T}-1)^2} l_A dE_A,
\eeq
where $ l_A $ is the mean free path.

@@

\begin{figure}[th]
	\includegraphics[width=9cm]{endtoend.png}
	\caption{A comparison of the final result for the excluded region with the result from Ref.~\cite{dent_etal12}. The excluded region is clearly significantly less restrictive, and opens up the region of parameter space for $\varepsilon \approx  1e-8-9 $.
\arz{Cameron, some similar comments as for Figure 1. Can we write numbers as "$10^{-9}$" rather 
than "1e-09" and so on? Can we make the axis labels and tick mark labels larger?}}
\label{fig:constraint}
\end{figure}

@@
